BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2016-12-26 Paper/BOAST.tex
jusqu a section 4

The paper is organized as follows. First in Section~\ref{sec:backg} we further motivate the need for metaprogramming frameworks.  Secondly in
Section~\ref{sec:laplace} we illustrate the optimization problems related to scientific computing with the Laplace equation. Thirdly we present our
solution, called BOAST, its architecture and its functionalities in Section~\ref{sec:boast}. Fourthly we apply BOAST to three use cases: the Laplace
equation motivating example, short multi-dimensional convolutions from the BigDFT quantum chemistry application, and the seismic wave propagation
simulator SPECFEM3D where 80 GPU computing kernels have been ported improving considerably the maintainability and providing a single code base for
OpenCL and CUDA kernel versions. Finally, before a conclusion and future works, we compare BOAST with other related works.

@@

The motivation of the BOAST project comes from the need of the HPC communities
which have full and large applications.  These applications are difficult to
port and maintain for emerging architectures. Especially since new
architectures are based on accelerators, longer vector units, or low power
processors.

%%%
\subsection{Evolution of HPC Architectures}

Evolution of HPC Architectures is rapid and also diverse: in the last 7 years
no less than 6 architectures have been number one in the
Top500~\cite{webtop500}:
\begin{itemize}
\item Intel Processor + Xeon Phi (Tianhe-2, 2013)
\item AMD Processor + NVIDIA GPU (Titan, 2012)
\item IBM BlueGene/Q (Sequoia, 2012)
\item Fujitsu SPARC64 (K computer, 2011)
\item Intel Processor + NVIDIA GPU (Tianhe-1, 2010)
\item AMD Processor (Jaguar, 2009)
\end{itemize}
Being able to efficiently use those architectures on such a small period is
challenging for scientists and application developers.

@@

\subsection{Scientific Computing Applications}

Scientific Computing Applications are usually developed by other science
researchers such as physicists, chemists or meteorologists. Those codes are
usually written in Fortran for historical and performance reasons. Codes can be
quite huge (several thousands lines of code [LOC]) with lots of
functionalities. Nonetheless, they are usually based on computing kernels.
Computing kernels are resource intensive and well-defined parts of a program
that usually work on precisely defined data. Those kernels represent the most
time-consuming part of an HPC application, and consequently, they are the prime
target for optimization.

Those applications are often developed by several individuals. Sometimes, some
of those developers only work a few months on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually the first available while
Fortran may sometimes come later.

@@

\subsection{How Should Computing Kernels be Written?}

@@

%%%
\section{Basic Example: OpenCL Laplace\label{sec:laplace}}

During one of the Mont-Blanc face to face meeting a talk on OpenCL optimization
on the Mali GPU was given. One of the case study was a Laplace filter~\cite{opencl_arm_training}.

This is a good example of an algorithm that is simple in its formulation but
can be complex to optimize because many different optimizations are available
and can be combined together. The correct combination of optimizations will
depend on the target architecture.

\subsection{The Laplace Filter}

Listing~\ref{lst:laplace_c_filter} shows a simple implementation of the Laplace
filter in C99. For clarity, boundary conditions have been omitted.

@@

\subsection{Possible Optimizations of the OpenCL Version}

@@

\paragraph{Vectorization}
The first proposed optimization involves computing five pixels instead of one
using the vectors available on the Mali architecture. The vectors have to be
kept in memory to ensure an efficient execution. Using vectors of 16 elements,
15 useful components are simultaneously loaded and can be used in computation.
This optimization yields speedups between 1.5 and 6 depending on the image
size.

@@

\subsection{Improving the Methodology}

@@

In the next Section we will present our answer to these issues: BOAST.  BOAST is
a framework dedicated to kernel description, optimization, regression testing
and autotuning.

\section{BOAST: Using Code Generation in Application Autotuning\label{sec:boast}}

@@

\section{Related Work}

>>>>

The paper is organized as follows. First in Section~\ref{sec:backg} we further motivate the need for metaprogramming frameworks.  Secondly in
Section~\ref{sec:laplace} we illustrate the optimization problems related to scientific computing with the Laplace equation. Thirdly we present our
solution, called BOAST, its architecture and its functionalities in Section~\ref{sec:boast}. Fourthly we apply BOAST to three use cases: the Laplace
equation motivating example, short multi-dimensional convolutions from the BigDFT quantum chemistry application, and the seismic wave propagation
simulator SPECFEM3D where 80 GPU computing kernels have been ported improving considerably the maintainability and providing a single code base for
OpenCL and CUDA kernel versions. Finally, before a conclusion and future works, we compare BOAST with other related works in
Section~\ref{sec:related}.

@@

The motivation of the BOAST project comes from the need of the HPC communities which have full and large applications.  These applications are
difficult to port and maintain for emerging architectures especially since new architectures are based on accelerators, longer vector units, or low
power processors.

%%% EVOLUTION OF HPC ARCHITECTURES %%%
\subsection{Evolution of HPC Architectures}

Evolution of HPC Architectures is rapid and also diverse. In the last seven years no less than 6 architectures have been number one in the
Top500~\cite{webtop500}:
\begin{itemize}
\item Intel Processor + Xeon Phi (Tianhe-2, 2013)
\item AMD Processor + NVIDIA GPU (Titan, 2012)
\item IBM BlueGene/Q (Sequoia, 2012)
\item Fujitsu SPARC64 (K computer, 2011)
\item Intel Processor + NVIDIA GPU (Tianhe-1, 2010)
\item AMD Processor (Jaguar, 2009)
\end{itemize}

Being able to efficiently use those architectures on such a small period is challenging for scientists and application developers.

@@

%%% SCIENTIFIC COMPUTING APPLICATIONS %%%
\subsection{Scientific Computing Applications}

Scientific Computing Applications are usually developed by other science researchers such as physicists, chemists or meteorologists. Those codes are
usually written in Fortran for historical and performance reasons. Codes can be quite huge (several thousands Lines of Code, LoC) with lots of
functionalities. Nonetheless, they are usually based on computing kernels.  Computing kernels are resource intensive and well-defined parts of a
program that usually work on precisely defined data. Those kernels represent the most time-consuming part of an HPC application, and consequently,
they are the prime target for optimizations.

Those applications are often developed by several individuals. Sometimes, some of those developers only work a few months on the
application. Maintaining optimized code written by someone else is quite a challenge. Several languages and programming paradigms can also be used in
a project and thus the maintainer must be knowledgeable in several areas of expertise. Portability problems can also be caused by the availability of
an optimizing compiler for a specific language and architecture. C compilers are usually first available while Fortran may sometimes come later.

@@

%%% HOW SHOULD COMPUTING KERNELS BE WRITTEN? %%%
\subsection{How Should Computing Kernels be Written?}

@@

%%% BASIC EXAMPLE: OPENCL LAPLACE %%%
\section{Basic Example: OpenCL Laplace\label{sec:laplace}}

During one of the Mont-Blanc face to face meeting a talk on OpenCL optimization on the Mali GPU was given. One of the case study was a Laplace
filter~\cite{opencl_arm_training}. This is a good example of an algorithm that is simple in its formulation but can be complex to optimize because
many different optimizations are available and can be combined together. The correct combination of optimizations will depend on the target
architecture.

%%% THE LAPLACE FILTER %%%
\subsection{The Laplace Filter}

Listing~\ref{lst:laplace_c_filter} shows a simple implementation of the Laplace filter in C99. For sake of clarity, boundary conditions have been
omitted.

@@

%%% POSSIBLE OPTIMIZATIONS OF THE OPENCL VERSION %%%
\subsection{Possible Optimizations of the OpenCL Version}

@@

%%%
\paragraph{Vectorization} The first proposed optimization involves computing five pixels instead of one using the vectors available on the Mali
architecture. The vectors have to be kept in memory to ensure an efficient execution. Using vectors of 16 elements, 15 useful components are
simultaneously loaded and can be used in computation.  This optimization yields speedups between 1.5 and 6 depending on the image size.

@@

%%% Improving the Methodology %%%
\subsection{Improving the Methodology}

@@

In the next Section we will present our answer to these issues: BOAST.  BOAST is a framework dedicated to kernel description, optimization, regression
testing, and autotuning.

%%% BOAST: USING CODE GENERATION IN APPLICATION AUTOTUNING %%%
\section{BOAST: Using Code Generation in Application Autotuning\label{sec:boast}}

@@

%%%
\section{Related Work\label{sec:related}}
